\fontfamily{phv}\selectfont{Causal Discovery methods for Effective Connectivity in Human Brains}

### 2014-01-16 docs/paper/paper final.tex
Wrote part on 'what data is needed'

\subsection{PC algorithm}
%Algorithm, assumptions and explanation [Ramon: standaard PC een naam geven en consistent gebruiken, bijv. standard PC?]
\paragraph{Description}
The PC algorithm is a so-called constraint based method, where solutions are found by incrementally leaving out or adding parts that do not violate a set of constraints.
Starting from a fully connected undirected graph, the PC algorithm forms structure by cutting away edges between conditionally independent variables.
In the second step, orientation rules are applied based on found separating sets and structure.
The original PC algorithm assumes causal sufficiency, excluding the existence of common confounders between nodes.
This seems unwanted when handling brain data, where one would expect a large amount of latent processes.
Also measuring methods are not yet advanced enough to measure detailed neuronal processing of the brain, thus skipping over details.
In the perspective of causal inference, these missing areas can also appear as confounders.
For the following experiments we have used the causal insufficient version of PC algorithm, modified to handle latent variable as described by Spirtes, Glymour and Scheines \cite[p.165-167]{spirtes2000}.
From now on, we will refer to this algorithm as \textit{standard PC}.
Pseudocode for this method is shown in algorithm \ref{alg:pc_alg}.

@@

%Before finding directed edges, the PC algorithm first finds undirected structure.
%For each two non-adjacent nodes $X$ and $Y$ standard PC finds a minimal separating set, which is the smallest set acting as a separating set.
%This is done by first assuming all nodes are connected, and then testing all pairs of nodes for independence.

@@

\paragraph{Multiple Separating Sets}
As will be explained in the results section, standard PC is a bit hasty in assigning v-structures, resulting in much unnecessary and perhaps unfaithful directionality.
The idea of CPC, to make a weaker assumption on graph faithfulness than standard PC is potential as less erroneous arrows are found.
Instead of doing additional tests as described in CPC, one could attempt to determine causal unfaithfulness with less dependency tests.
An adaptation to PC algorithm that would not increase the order of complexity is once a separating set is found to test all other vertex subsets of the same order as well.
This finds all separating sets of the lowest order without increasing complexity.
Orienting an unshielded triple $\left<A,B,C\right>$ requires $B$ not to occur in any separating set of $(A,C)$.
By having more than one separating set, this constraint is strengthened in case of noisy data.

% Unfaithfulness test
One could use these additional separating sets to test oriented unshielded triples for unfaithfulness, similar to CPC.
Apart from serving as a sanity check, this could serve as the basis of a repair algorithm for errors made by PC.

% Explicit test
A more explicit way to decrease faulty orientation of an unshielded triple $\left<A,B,C\right>$ is to check whether adding $B$ to a separating set of $(A,C)$ makes the pair conditionally dependent.
If so, the graph might become more faithful to its modeled distribution; If standard PC might orient a V-structure, whereas this condition does not hold, then the concluded V-structure might be erroneous.
For faithful graphs, this extra condition should not change the result, as it will always hold.

@@

\subsection{Experimental validation}
\paragraph{Experimental pipeline}{sec:pipeline}
%- What data is needed
%- Apply structure PC on resting-state functional time-series
%- Apply directional PC
Both standard PC and EMS-pc have been used to find brain connectivity.
The data which is used is functional time-series data, and it is the same as the data used in Hinne et al.\cite{hinne2013}, as described in Section~\ref{sec:data}.
This data is obtained from six subject.
From this data, covariance between regions can be determined, so that regions can be tested for \textit{d}-separation with the Fisher's conditional independence test.
This is required to use the PC algorithm.
First, the strucural part of the PC algorithm will be applied on the data.
This will result in a graph describing undirected connections between regions.
There is no difference between standard PC and EMS-PC in this result.
After that, the second part of the PC algorithm is run to orient edges in this graph.
This is different for standard PC and EMS-PC and a result will therefore be obtained for both algorithms.

@@

\subsection{Data acquisition}\label{sec:data}
%Subject acquisition and instruction
%Used equipment and specs
%Experimental setup
%Scanner settings
%Preprocessing
%- Software
%- Scripts
%Resulting data specs
%TODO niet helemaal happy, hoe zeg je dit zonder steeds "as described by Hinne et al." te zeggen? "This research" kan ook naar ons eigen werk verwijzen en is verwarrend, het moet duidelijk zijn dat dit alles niet door ons gedaan is.
We performed our analysis on the same data as treated by Hinne et al. in the aforementioned Bayesian structural analysis \cite{hinne2013, hinne2013structfunc}.
This article describes how resting-state functional data of six subjects was measured at the Donders Centre for Cognitive Neuroimaging at the Radboud University.
This functional data is divided in 116 nodes and 1029 time-series per subject.
Additionally, diffusion-weighted images of these subjects have been obtained.
Based on these DWI-measurements, a structural connectivity matrix has been inferred with a technique described by Hinne et al.
We used the six resulting resting-state functional time series data and connectivity matrices from the Bayesian analysis as pre-processed by Hinne et al.

>>>>

\subsection{PC algorithm}
%Algorithm, assumptions and explanation
\paragraph{Description}
The PC algorithm is a so-called constraint based method, where solutions are found by incrementally leaving out or adding parts that do not violate a set of constraints.
Starting from a fully connected undirected graph, the PC algorithm forms structure by cutting away edges between conditionally independent variables.
In the second step, orientation rules are applied based on found separating sets and structure.
The original PC algorithm assumes causal sufficiency, excluding the existence of common confounders between nodes.
This seems unwanted when handling brain data, where one would expect a large amount of latent processes.
Also measuring methods are not yet advanced enough to measure detailed neuronal processing of the brain, thus skipping over details.
In the perspective of causal inference, these missing areas can also appear as confounders.
For the following experiments we have used the causal insufficient version of PC algorithm, modified to handle latent variable as described by Spirtes, Glymour and Scheines \cite[p.165-167]{spirtes2000}.
From now on, we will refer to this algorithm as \textit{standard PC}.
Pseudocode for this method is shown in algorithm \ref{alg:pc_alg}.

@@

@@

\paragraph{Multiple Separating Sets}
As can be seen in the result in the next section, standard PC is a bit hasty in assigning v-structures, resulting in much unnecessary and perhaps unfaithful directionality.
The idea of CPC, to rely less on graph faithfulness than standard PC is potential as less erroneous arrows are assigned in theory.
Instead of doing additional tests as described in CPC, one could attempt to determine causal unfaithfulness with less dependency tests.
An adaptation to PC algorithm that would not increase the order of complexity is once a separating set is found to test all other vertex subsets of the same order as well.
This finds all separating sets of the lowest order without increasing complexity.
Orienting an unshielded triple $\left<A,B,C\right>$ requires $B$ not to occur in any separating set of $(A,C)$.
By having more than one separating set, this constraint is strengthened in case of noisy data.
% Unfaithfulness test
One could use these additional separating sets to test oriented unshielded triples for unfaithfulness, similar to CPC.
Apart from serving as a sanity check, this could serve as the basis of a repair algorithm for errors made by PC.

% Explicit test
A more \textit{explicit} way to decrease faulty orientation of an unshielded triple $\left<A,B,C\right>$ is to check whether adding $B$ to a separating set of $(A,C)$ makes the pair conditionally dependent.
If so, the graph might become more faithful to its modeled distribution; If standard PC might orient a V-structure, whereas this condition does not hold, then the concluded V-structure might be erroneous.
For faithful graphs, this extra condition should not change the result, as it will always hold.

@@

\subsection{Experimental validation}
\paragraph{Experimental pipeline}{sec:pipeline}
%- What data is needed
PC algorithm requires observational data where each instance is a vector of single observations per variable.
We can model functional resting-state time-series as coming from some stable underlying mechanism that is largely time-independent.
Resting-state data can be intuitively thought of being more time independent than task-influenced data due to the emergent character of activity pattens that occur when in resting-state.
Nevertheless it is most likely not a fully correct modelling assumption, though enables straightforward application of standard and EMS-PC to functional data without the problem of HRF.

%- Apply structure PC on resting-state functional time-series
%- Apply directional PC
Both standard PC and EMS-pc have been used to find brain connectivity.
The data which is used is functional time-series data, and it is the same as the data used in Hinne et al.\cite{hinne2013}, as described in Section~\ref{sec:data}.
This data is obtained from six subject.
From this data, covariance between regions can be determined, so that regions can be tested for \textit{d}-separation with the Fisher's conditional independence test.
This is required to use the PC algorithm.
First, the strucural part of the PC algorithm will be applied on the data.
This will result in a graph describing undirected connections between regions.
There is no difference between standard PC and EMS-PC in this result.
After that, the second part of the PC algorithm is run to orient edges in this graph.
This is different for standard PC and EMS-PC and a result will therefore be obtained for both algorithms.

@@

% TODO: Move to appendix?
\subsection{Data acquisition}\label{sec:data}
%Subject acquisition and instruction
%Used equipment and specs
%Experimental setup
%Scanner settings
%Preprocessing
%- Software
%- Scripts
%Resulting data specs
%TODO niet helemaal happy, hoe zeg je dit zonder steeds "as described by Hinne et al." te zeggen? "This research" kan ook naar ons eigen werk verwijzen en is verwarrend, het moet duidelijk zijn dat dit alles niet door ons gedaan is.
We performed our analysis on the same data as treated by Hinne et al. in the aforementioned Bayesian structural analysis \cite{hinne2013, hinne2013structfunc}.
This article describes how resting-state functional data of six subjects was measured at the Donders Centre for Cognitive Neuroimaging at the Radboud University.
This functional data is divided in 116 nodes and 1029 time-series per subject.
Additionally, diffusion-weighted images of these subjects have been obtained.
Based on these DWI-measurements, a structural connectivity matrix has been inferred with a technique described by Hinne et al.
We used the six resulting resting-state functional time series data and connectivity matrices from the Bayesian analysis as pre-processed by Hinne et al.
